---
title: "Patescibacteria phylogenies"
author: "Rob Lanfear"
date: "`r Sys.Date()`"
output:
  html_document:
    css: ../../style.css
    toc: yes
    toc_depth: 4
    theme: cosmo
    number_sections: no
    toc_float: yes
    highlight: pygments
    fig_width: 8
    fig_height: 4
  word_document:
    toc: yes
    toc_depth: '4'
  pdf_document:
    toc: yes
    toc_depth: '4'
    pandoc_args: --listings
    includes:
      in_header: preamble.tex
editor_options: 
  chunk_output_type: console
---


# Intro 

Let's look at the trees and models we've built for the patescibacteria.

## Datasets

We have 4 datasets: r202, r207, r214, r220. For each, we have the GTDB tree estimated in FastTree, and then my tree estimated with a GTR20+C60 PMSF model in IQ-TREE.

The datasets have increasing numbers of taxa for patescibacteria:

* r202: 1601 taxa
* r207: 2485 taxa
* r214: 3374 taxa
* r220: 4581 taxa

## tree comparisons

For r202 I did two analyses with identical scripts, to see if the method is repeatable. 

Let's load the FastTree and the IQ-TREE tree for each dataset

```{r}
library(phangorn)
setwd("/Users/roblanfear/Documents/github/bacterial-phylo/phlya/patescibacteria_analysis")
r202a_ft <- unroot(read.tree("r202/subtree.nwk"))
r202a_iq <- read.tree("r202/06_PMSF_tree_ufboot.treefile")
r202b_ft <- unroot(read.tree("r202b/subtree.nwk"))
r202b_iq <- read.tree("r202b/06_PMSF_tree_ufboot.treefile")
r207_ft <- unroot(read.tree("r207/subtree.nwk"))
r207_iq <- read.tree("r207/06_PMSF_tree_ufboot.treefile")
r214_ft <- unroot(read.tree("r214/subtree.nwk"))
r214_iq <- read.tree("r214/06_PMSF_tree_ufboot.treefile")
r220_ft <- unroot(read.tree("r220/subtree.nwk"))
r220_iq <- read.tree("r220/06_PMSF_tree_ufboot.treefile")

# consensus trees - often better in ML terms
r202a_ct <- read.tree("r202/06_PMSF_tree_ufboot.contree")
r202b_ct <- read.tree("r202b/06_PMSF_tree_ufboot.contree")
r207_ct <- read.tree("r207/06_PMSF_tree_ufboot.contree")
r214_ct <- read.tree("r214/06_PMSF_tree_ufboot.contree")
r220_ct <- read.tree("r220/06_PMSF_tree_ufboot.contree")
```

A few simple comparisons

```{r}
library(tidyverse)
library(ggdendro)
library(viridis)
library(phangorn)
library(patchwork)

# Load and prepare trees
r202a_ft <- unroot(read.tree("r202/subtree.nwk"))
r202a_iq <- read.tree("r202/06_PMSF_tree_ufboot.treefile")
r202b_ft <- unroot(read.tree("r202b/subtree.nwk"))
r202b_iq <- read.tree("r202b/06_PMSF_tree_ufboot.treefile")
r207_ft <- unroot(read.tree("r207/subtree.nwk"))
r207_iq <- read.tree("r207/06_PMSF_tree_ufboot.treefile")
r214_ft <- unroot(read.tree("r214/subtree.nwk"))
r214_iq <- read.tree("r214/06_PMSF_tree_ufboot.treefile")
r220_ft <- unroot(read.tree("r220/subtree.nwk"))
r220_iq <- read.tree("r220/06_PMSF_tree_ufboot.treefile")

# consensus trees
r202a_ct <- read.tree("r202/06_PMSF_tree_ufboot.contree")
r202b_ct <- read.tree("r202b/06_PMSF_tree_ufboot.contree")
r207_ct <- read.tree("r207/06_PMSF_tree_ufboot.contree")
r214_ct <- read.tree("r214/06_PMSF_tree_ufboot.contree")
r220_ct <- read.tree("r220/06_PMSF_tree_ufboot.contree")

# Find common tips
common_tips <- Reduce(intersect, list(
  r202a_ft$tip.label, r202a_iq$tip.label, r202b_ft$tip.label, r202b_iq$tip.label,
  r207_ft$tip.label, r207_iq$tip.label, r214_ft$tip.label, r214_iq$tip.label,
  r220_ft$tip.label, r220_iq$tip.label, r202a_ct$tip.label, r202b_ct$tip.label,
  r207_ct$tip.label, r214_ct$tip.label, r220_ct$tip.label
))

# Prune trees to common tips
prune_tree <- function(tree) drop.tip(tree, setdiff(tree$tip.label, common_tips))
all_trees <- list(r202a_ft, r202a_iq, r202b_ft, r202b_iq,
                  r207_ft, r207_iq, r214_ft, r214_iq,
                  r220_ft, r220_iq, r202a_ct, r202b_ct,
                  r207_ct, r214_ct, r220_ct)
pruned_trees <- lapply(all_trees, prune_tree)

# Set tree names
names(pruned_trees) <- c("r202a_ft", "r202a_iq", "r202b_ft", "r202b_iq",
                         "r207_ft", "r207_iq", "r214_ft", "r214_iq",
                         "r220_ft", "r220_iq", "r202a_ct", "r202b_ct",
                         "r207_ct", "r214_ct", "r220_ct")

class(pruned_trees) <- "multiPhylo"

# Calculate normalized wRF distances
wRFn <- path.dist(pruned_trees, check.labels = TRUE)

# Function to convert lower triangle to full symmetric matrix
full_matrix <- function(lower_tri_matrix) {
  full <- as.matrix(lower_tri_matrix)
  full[upper.tri(full)] <- t(full)[upper.tri(full)]
  diag(full) <- 0
  return(full)
}

# Convert to full matrix
wRFn_full <- full_matrix(wRFn)

# Hierarchical clustering
hc <- hclust(as.dist(wRFn_full))
dendro_order <- hc$labels[hc$order]

# Prepare data for heatmap
plot_data <- as.data.frame(wRFn_full) %>% 
  rownames_to_column('Tree1') %>%
  pivot_longer(-Tree1, names_to = "Tree2", values_to = "Distance") %>%
  mutate(Tree1 = factor(Tree1, levels = dendro_order),
         Tree2 = factor(Tree2, levels = dendro_order))

# Dendrogram plot
 dendro_data <- ggdendro::dendro_data(as.dendrogram(hc))
 dendrogram_plot <- ggplot() +
   geom_segment(data = dendro_data$segments,
                aes(x = x, y = y, xend = xend, yend = yend)) +
   scale_x_continuous(breaks = seq_along(dendro_data$labels$label),
                      labels = dendro_data$labels$label, expand = c(0, 0.5)) +
   coord_flip() +
   theme_minimal() +
   labs(y = "Height", x = NULL) +
   theme(axis.title.x = element_blank(), axis.text.x = element_blank(), axis.ticks.x = element_blank())

# Heatmap plot
heatmap_plot <- ggplot(plot_data, aes(x = Tree1, y = Tree2, fill = Distance)) +
  geom_tile() +
  scale_fill_viridis(option = "viridis") +
  labs(title = "Path Distances between trees", x = "", y = "", subtitle = "The trees from IQ-TREE are similar to each other, and to the r220 GTDB tree. The other GTDB trees are different...") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(hjust = 0.5, face = "bold"))

# Combine plots
dendrogram_plot + heatmap_plot + plot_layout(widths = c(1, 4))
```

